---
title: "Darwin Core mapping"
subtitle: "For: Checklist of non-native freshwater fishes in Flanders, Belgium"
author:
- Lien Reyserhove
- Peter Desmet
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-fishes-checklist/blob/master/src/dwc_mapping.Rmd).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)      # To do data science
library(magrittr)       # To use %<>% pipes
library(here)           # To find files
library(janitor)        # To clean input data
library(readxl)         # To read Excel files
library(digest)         # To generate hashes
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/ExotischeVissenVlaanderen2016.xlsx"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_vernacular_file = "../data/processed/vernacularname.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
dwc_description_file = "../data/processed/description.csv"
dwc_profile_file = "../data/processed/speciesprofile.csv"
```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_excel(raw_data_file, sheet = "Checklist", na = "NA")
```

Clean the data somewhat: remove empty rows if present

```{r}
raw_data %<>%
  remove_empty("rows") %>%    # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable:

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("alien-fishes-checklist", "taxon", vdigest(latin_name, algo="md5"), sep = ":"))
```

## Further pre-processing:

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping
 
Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "INBO")

```

### accessRights

```{r}
taxon %<>% mutate(accessRights = "http://www.inbo.be/en/norms-for-data-use")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "https://doi.org/10.15468/xvuzfh")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Checklist of non-native freshwater fishes in Flanders, Belgium")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_latin_name)
```

Verify that `scientificName` contains unique values:

```{r}
any(duplicated(taxon$scientificName)) # Should be FALSE
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = "Animalia")
```

### taxonRank

All taxa are species:

```{r}
taxon %<>% mutate(taxonRank = "species")
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(nomenclaturalCode = "ICZN")
```

## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("raw_"))
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write_csv(taxon, dwc_taxon_file, na = "")
```

# Create vernacular names extension

```{r start_vernacular_names}
vernacular_names <- raw_data
```

## Term mapping

Map the data to [Vernacular Names](http://rs.gbif.org/extension/gbif/1.0/vernacularname.xml).

### taxonID

```{r}
vernacular_names %<>% mutate(taxonID = raw_taxon_id)
```

### vernacularName

Vernacular names are available in two languages: English (`raw_common_name`) and Dutch (`raw_nederlandse_naam`). We will gather these columns to generate a single column containing the vernacular name (`vernacularName`) and an additional column with the language (`language`):

```{r}
vernacular_names %<>%
  gather(key = language, value = vernacularName, raw_common_name, raw_nederlandse_naam, na.rm = TRUE, convert = TRUE) %>%
  select(-language, everything()) # Move language column to the end
```

```{r}
vernacular_names %<>% arrange(taxonID)
```

### language

This column currently contains the original column name, which we will recode to the ISO 639-1 language code:

```{r}
vernacular_names %<>% mutate(language = recode(language,
  "raw_common_name" = "en",
  "raw_nederlandse_naam" = "nl"
))
```

## Post-processing

Remove the original columns: 

```{r}
vernacular_names %<>% select(
  -starts_with("raw"))

```

Preview data:

```{r}
vernacular_names %>% head()
```

Save to CSV:

```{r}
write_csv(vernacular_names, dwc_vernacular_file, na = "")
```

# Create distribution extension

```{r start_distribution}
distribution <- raw_data
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(taxonID = raw_taxon_id)
```

### locationID

```{r}
distribution %<>% mutate(locationID = "ISO_3166-2:BE-VLG")
```

### locality

```{r}
distribution %<>% mutate(locality = "Flemish Region")
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = "BE")
```

### occurrenceStatus

```{r}
distribution %<>% mutate(occurrenceStatus = "present")
```

### establishmentMeans

We use the [GBIF controlled vocabulary](http://rs.gbif.org/vocabulary/gbif/establishment_means.xml) for this term. For this dataset, all species are `introduced` (= alien):

```{r}
distribution %<>% mutate(establishmentMeans = "introduced")
```

### degreeOfEstablishment

Invasion stage information (e.g. `casual`) can be found in `raw_status`. 

Create separate dataframe:

Show unique values:

```{r}
distribution %>%
  distinct(raw_status) %>%
  arrange(raw_status)
```

* `A` and `A*` in the raw data file stand for `acclimatized` and refers to the status where the alien species has overcome the geographical barrier, the captivity/cultivation barrier and the survival barier, but where it fails to establish because individuals in the population fail to reproduce. This includes species (like Nile tilapia) in secluded warmer water of power plants. In Blackburn et al. (2011), this corresponds to the term `casual`.
* `N` in the raw data file stands for `naturalized`. We decided **not to use** this term, because often, there's no sensible criterium to distinguish between casual/naturalised and naturalised/established. Thus, here, we consider all `naturalized` species to be `established`.

```{r}
distribution %<>% mutate(degreeOfEstablishment = case_when(
  raw_status == "A" ~ "casual",
  raw_status == "A*" ~ "casual",
  raw_status == "N" ~ "established"
))
```

### pathway

Pathway information can be found in `raw_pathway_s`. Show unique values:

```{r}
distribution %>%
  distinct(raw_pathway_s) %>%
  arrange(raw_pathway_s)
```

`raw_pathway_s` contains multiple values (currently not more than 2), so we separate it in 2 columns:

```{r}
distribution %<>% separate(
  raw_pathway_s,
  into = c("pathway_1", "pathway_2"),
  sep = ", ",
  remove = FALSE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
distribution %<>% gather(
  key, value,
  pathway_1, pathway_2,
  na.rm = TRUE, # Also removes records for which there is no pathway_1
  convert = FALSE
)
```

This is a manually created overview of the abbreviations used in this dataset, their interpretation and the mapping to the CBD vocabulary:

```{r, echo = FALSE}
as.data.frame(matrix(data = c(
  "AM", "Active Migration", "corridor_water",
  "AN", "Angling/bait fish","escape_food_bait",
  "AQ", "Aquaculture", "escape_aquaculture",
  "BC", "Biological Control", "release_biologica_control",
  "BW", "Ballast Water", "stowaway_ballast_water",
  "OR", "Ornamental", "escape_ornamental",
  "UN", "Unintentional", "Unintentional"),
  nrow = 7, ncol = 3, byrow = TRUE,
  dimnames = list(c(1:7), c("abbreviation", "full name", "CBD mapping"))
))
```

Map values:

```{r}
distribution %<>% mutate(pathway = recode(value,
  "AM" = "corridor",
  "AN" = "liveFoodBait",
  "AQ" = "aquacultureMariculture",
  "BC" = "biologicalControl",
  "BW" = "ballastWater",
  "OR" = "pet",
  "UN" = "unintentional"
))
```

Show mapped values:

```{r}
distribution %>%
  select(value, pathway) %>%
  group_by_all() %>%
  summarize(records = n()) %>%
  arrange(value) 
```

### eventDate

The distribution information applies to a certain date range, which we will express here as an ISO 8601 date `yyyy/yyyy` (`start_year/end_year`).

The `start_year` is the year of first observation:

```{r}
distribution %<>% mutate(start_year = raw_introduction)
```

Some `start_year` values are quite broad. We change these values to the first year for which an observation is plausible:

```{r}
distribution %<>% mutate(start_year = recode(start_year,
  "20xx" = "2000",
  "17th c." = "1601",
  "1980s" = "1980",
  "13th c." = "1201"
))
```

As there is no `end_year` information, we'll leave this field empty:

```{r}
distribution %<>% mutate(end_year = "")
```

Create `eventDate` by combining `start_year/end_year`:

```{r}
distribution %<>% mutate(eventDate = paste(start_year, end_year, sep = "/"))
```

Compare formatted dates with original dates in `raw_introduction`:

```{r}
distribution %>%
  distinct(raw_introduction, eventDate) %>%
  arrange(raw_introduction)
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("raw_"), -start_year, -end_year, -key, -value)
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write_csv(distribution, dwc_distribution_file, na = "")
```

# Create description extension

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

Create `description`:

```{r}
description <- raw_data
```

In the description extension we want to include the native range of the species.
A single taxon can have multiple native ranges of the same type, expressed as multiple rows in the description extension. The mapping will be as following:

- `type` = `native range`
- `description` = the specific content of the field `raw_origin`
- `language` = `en`

## Term mapping

### taxonID

```{r}
description %<>% mutate(taxonID = raw_taxon_id)
```

### description

Show unique values:

```{r}
description %>%
  distinct(raw_origin) %>%
  arrange(raw_origin)
```
`raw_origin` contains multiple values (currently not more than 2), so we separate it on `" or "` in 2 columns:

```{r}
description %<>% separate(raw_origin,
  into = c("native_range_1", "native_range_2"),
  sep = " or ",
  remove = FALSE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
description %<>% gather(
  key, value,
  native_range_1, native_range_2,
  na.rm = TRUE, # Also removes records for which there is no native_range_1
  convert = FALSE
)
```

Map values:

```{r}
description %<>% mutate(mapped_value = recode(value,
  "AS" = "Asia",
  "EE" = "Eastern Europe",
  "AFR" = "Africa"))
```

Show mapped values:

```{r}
description %>%
  select(value, mapped_value) %>%
  group_by_all() %>%
  summarize(records = n()) %>%
  arrange(value) 
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
description %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

### type
 
Create a `type` field to indicate the type of description:

```{r}
description %<>% mutate(type = "native range")
```

### language

```{r}
description %<>% mutate(language = "en")
```

## Post-processing

Remove the original columns:

```{r}
description %<>% select(-starts_with("raw_"))
```

Move `taxonID` to the first position:

```{r}
description %<>% select(taxonID, everything())
```

Sort on `taxonID` to group description information per taxon:

```{r}
description %<>% arrange(taxonID)
```

Preview data:

```{r}
description %>% head(10)
```

Save to CSV:

```{r}
write_csv(description, dwc_description_file, na = "")
```

# Create species profile extension

In this extension will express broad habitat characteristics of the species.

```{r start_species_profile}
species_profile <- raw_data
```

Habitat information can be found in `raw_habitat`. All taxa are considered to be freshwater fishes. 

## Term mapping

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

### taxonID

```{r}
species_profile %<>% mutate(taxonID = raw_taxon_id)
```

### isMarine
```{r}
species_profile %<>% mutate(isMarine = FALSE)
```

### isFreshwater

```{r}
species_profile %<>% mutate(isFreshwater = TRUE)
```

### isTerrestrial

```{r}
species_profile %<>% mutate(isTerrestrial = "FALSE")
```

## Post-processing

Remove the original columns:

```{r}
species_profile %<>% select(-starts_with("raw_"))
```

Sort on `taxonID`:

```{r}
species_profile %<>% arrange(taxonID)
```

Preview data:

```{r}
species_profile %>% head()
```

Save to CSV:

```{r}
write_csv(species_profile, dwc_profile_file, na = "")
```
