---
title: "Darwin Core mapping"
subtitle: "For: Checklist of non-native freshwater fishes in Flanders, Belgium"
author: 
- Lien Reyserhove
- Peter Desmet
- Sanne Govaert
date: "`r Sys.Date()`"
output:
  html_document:
---

This script prepares checklist data for publication on GBIF. The source file for this document can be found [here](https://github.com/trias-project/alien-fishes-checklist/blob/master/src/dwc_mapping.Rmd).
It downloads the checklist data from [this Google Spreadsheet](https://docs.google.com/spreadsheets/d/1ViuIWMhDJ-drs2PkunC4DOrNSaNwleuWR1imUU4KzVA/edit#gid=315843027), where these data are maintained. The script transforms the data to a [Darwin Core checklist](https://www.gbif.org/dataset-classes) that can be harvested by [GBIF](http://www.gbif.org). It was developed for the [TrIAS project](http://trias-project.be).

```{r setup, include = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Load libraries:

```{r}
library(tidyverse)
library(magrittr)
library(janitor)
library(digest)
library(here)
```

# Read source data

Read the relevant worksheet (published as csv) and Copy the source data to the repository to keep track of changes.

```{r read_source_data}
raw_data <-read_csv("https://docs.google.com/spreadsheets/d/e/2PACX-1vSXWde261c45tj0pIMeESy15M3BMogW-1Q-92a4m7A8aKtztHi4JPKEJHDSjEHYHGjQ6de1z27avnOs/pub?gid=0&single=true&output=csv", na = "NA")
readr::write_csv(input_data, here("data", "raw", "alien_fisches_checklist_dump.csv"), na = "")
```
# Preprocessing: tidy data and add taxon ID's

To link taxa with information in the extension(s), each taxon needs a unique and relatively stable `taxonID`. We have created one in the form of `dataset_shortname:taxon:hash`, where `hash` is unique code based on scientific name and kingdom. Once this is created, it is added to the source data. 

```{r}
input_data <-
  raw_data %>%
  remove_empty("rows") %>%
  clean_names() %>%
  mutate(
    taxon_id = paste(
      "alien-fishes-checklist",
      "taxon",
      .data$taxon_id_hash,
      sep = ":"
    )
  )
```

# Darwin Core mapping

## Taxon core

Create a dataframe with unique taxa only (ignoring multiple distribution rows). Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).

```{r taxon_core}
taxon <-
  input_data %>%
  dplyr::distinct(taxon_id, .keep_all = TRUE) %>%
  dplyr::mutate(
    language = "en",
    license = "http://creativecommons.org/publicdomain/zero/1.0/",
    rightsHolder = "INBO",
    accessRights = "https://www.inbo.be/en/norms-data-use", 
    datasetID = "https://doi.org/10.15468/xvuzfh",
    institutionCode = "INBO",
    datasetName = "Checklist of non-native freshwater fishes in Flanders, Belgium", 
    taxonID = taxon_id,
    scientificName = scientific_name,
    kingdom = kingdom,
    phylum = phylum,
    order = order,
    family = family,
    genus = genus,
    taxonRank = taxon_rank,
    nomenclaturalCode = nomenclatural_code,
    .keep = "none"
  ) %>% 
 dplyr::arrange(taxonID) %>%
 dplyr::select(
   "language", "license", "rightsHolder", "accessRights", "datasetID",
   "institutionCode", "datasetName", "taxonID", "scientificName", "kingdom",
   "phylum", "order", "family", "genus", "taxonRank", "nomenclaturalCode"
 )
```

# Create vernacular names extension

Map the data to [Vernacular Names](http://rs.gbif.org/extension/gbif/1.0/vernacularname.xml).
Vernacular names are available in two languages: English (`raw_common_name`) and Dutch (`raw_nederlandse_naam`). We will gather these columns to generate a single column containing the vernacular name (`vernacularName`) and an additional column with the language (`language`):

```{r vernacular_names}
vernacular_names <- 
  input_data %>% 
  tidyr::gather(key = language, value = vernacularName, common_name, dutch_name, na.rm = TRUE, convert = TRUE) %>%
  dplyr::mutate(
    taxonID = taxon_id,
    language = 
      dplyr::recode(
        language,
        "common_name" = "en",
        "dutch_name" = "nl"
        ) 
    ) %>% 
  dplyr::select(taxonID, vernacularName, language) %>% 
  dplyr::arrange(taxonID)
```

# Create distribution extension

Create a dataframe with all data (including multiple distributions). Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

Information for `eventDate` is contained in `date_first_observation` and `date_last_observation`, which we will express here in an ISO 8601 date format `yyyy/yyyy` (`start_date/end_date`).

Not all cells for `date_first_observation` (DFO) and/or `date_last_observation` (DLO) are populated. So, we used the following rules for those records: 

***case 1.*** If `DFO` is empty and `DLO` is empty, `eventDate` is `NA` 
***case 2.***  If `DFO` is empty and `DLO` is not empty: eventDate = `/DLO`
***case 3.*** If `DFO` is not empty and `DLO` is empty, eventDate is `DFO/`

```{r distribution}
# dwc mapping
distribution <-
  input_data %>%
  # pathway
  pivot_longer(
    names_to = "key",
    values_to = "pathway",
    starts_with("introduction_pathway"),
    values_drop_na = FALSE) %>%
  filter( # keep NA value for species with no pathway provided
    !is.na(pathway) |
      (is.na(pathway) & key == "introduction_pathway_1")
    ) %>%
  # other terms
  dplyr::mutate(
    taxonID = taxon_id,
    locationID = dplyr::case_when(
      location == "Flanders" ~ "ISO_3166-2:BE-VLG",
      location == "Wallonia" ~ "ISO_3166-2:BE-WAL",
      location == "Brussels" ~ "ISO_3166-2:BE-BRU"
    ),
    locality = dplyr::case_when(
      location == "Flanders" ~ "Flemish Region",
      location == "Wallonia" ~ "Walloon Region",
      location == "Brussels" ~ "Brussels-Capital Region"
    ),
    countryCode = "BE",
    occurrenceStatus = occurrence_status,
    establishmentMeans = "introduced",
    degreeOfEstablishment = degree_of_establishment,
    pathway = pathway,
    eventDate = case_when(
      is.na(date_first_observation) & is.na(date_last_observation) ~ NA,
      is.na(date_first_observation) ~ paste0("/", date_last_observation),
      is.na(date_last_observation) ~ paste0(date_first_observation, "/"),
      !is.na(date_first_observation) & !is.na(date_last_observation) ~ 
        paste(date_first_observation, date_last_observation, sep = "/")
    ),
    source = source,
    occurrenceRemarks = remarks,
    .keep = "none") %>%
  select(
    "taxonID", "locationID", "locality", "countryCode", "occurrenceStatus",
    "establishmentMeans", "degreeOfEstablishment", "pathway", 
    "eventDate", "source", "occurrenceRemarks"
  ) 
```

# Create description extension

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).
`raw_origin` contains multiple values (currently not more than 2), so we separate it on `" or "` in 2 columns.

```{r description}
description <- 
  input_data %>% 
  dplyr::filter(!is.na(native_range)) %>% 
  dplyr::mutate(
    taxonID = taxon_id,
    description = strsplit(as.character(native_range), " or "),
    type = "native range",
    language = "en",
    .keep = "none"
  ) %>% 
  tidyr::unnest(description) %>% 
  dplyr::arrange(taxonID) 
```

## Species profile extension

In this extension we will express broad habitat characteristics of the species (e.g. `isTerrestrial`).

Create a dataframe with unique taxa only (ignoring multiple distribution rows).
Only keep records for which `terrestrial`, `marine` and `freshwater` is not empty.

Map the data to [Species Profile](http://rs.gbif.org/extension/gbif/1.0/speciesprofile.xml).

```{r}
species_profile <-
  input_data %>%
  distinct(taxon_id, .keep_all = TRUE) %>% 
  filter(
    !is.na(terrestrial) |
      !is.na(marine) |
      !is.na(freshwater)
  ) %>% 
  mutate(
    .keep = "none",
    taxonID = taxon_id,
    isMarine = marine,
    isFreshwater = freshwater,
    isTerrestrial = terrestrial
  ) %>% 
  arrange(taxonID)
```

# Save to CSV:

```{r}
readr::write_csv(taxon, file = here::here("data", "processed", "taxon.csv"), na = "")
readr::write_csv(vernacular_names, file = here::here("data", "processed", "vernacularname.csv"), na = "")
readr::write_csv(distribution, file = here::here("data", "processed", "distribution.csv"), na = "")
readr::write_csv(description, file = here::here("data", "processed", "description.csv"), na = "")
readr::write_csv(species_profile, file = here::here("data", "processed", "speciesprofile.csv"), na = "")
```

