---
title: "Darwin Core mapping"
subtitle: "For: Checklist of non-native freshwater fishes in Flanders, Belgium"
author:
- Lien Reyserhove
- Peter Desmet
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    number_sections: true
---

This document describes how we map the checklist data to Darwin Core. The source file for this document can be found [here](https://github.com/trias-project/alien-fishes-checklist/blob/master/src/dwc_mapping.Rmd).

# Setup

```{r setup, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
```

Set locale (so we use UTF-8 character encoding):

```{r}
# This works on Mac OS X, might not work on other OS
Sys.setlocale("LC_CTYPE", "en_US.UTF-8")
```

Load libraries:

```{r}
library(tidyverse) # To transform data
library(magrittr)  # For %<>% pipes
library(janitor)   # To clean input data
library(readxl)    # To read Excel files
library(stringr)   # to perform string operations
library(digest)    # To generate hashes
```

Set file paths (all paths should be relative to this script):
 
```{r}
# Raw files:
raw_data_file = "../data/raw/ExotischeVissenVlaanderen2016.xlsx"

# Processed files:
dwc_taxon_file = "../data/processed/taxon.csv"
dwc_vernacular_file = "../data/processed/vernacularname.csv"
dwc_distribution_file = "../data/processed/distribution.csv"
dwc_description_file = "../data/processed/description.csv"
```

# Read and pre-process raw data

Create a data frame `raw_data` from the source data:

```{r}
# Read the source data:
raw_data <- read_excel(raw_data_file, sheet = "Checklist", na = "NA")

# Clean the data somewhat: remove empty rows if present
raw_data %<>%
  remove_empty_rows() %>%     # Remove empty rows
  clean_names()               # Have sensible (lowercase) column names
```

## Generate taxonID

To uniquely identify a taxon in the taxon core and reference taxa in the extensions, we need a `taxonID`. Since we need it in all generated files, we generate it here in the raw data frame. It is a combination of `dataset-shortname:taxon:` and a hash based on the scientific name. As long as the scientific name doesn't change, the ID will be stable:

```{r}
# Vectorize the digest function (The digest() function isn't vectorized. So if you pass in a vector, you get one value for the whole vector rather than a digest for each element of the vector):
vdigest <- Vectorize(digest)

# Generate taxonID:
raw_data %<>% mutate(taxon_id = paste("alien-fishes-checklist", "taxon", vdigest(latin_name, algo="md5"), sep = ":"))
```

## Further pre-processing:

Add prefix `raw_` to all column names to avoid name clashes with Darwin Core terms:

```{r}
colnames(raw_data) <- paste0("raw_", colnames(raw_data))
```

Preview data:

```{r}
raw_data %>% head()
```

# Create taxon core

```{r start_taxon}
taxon <- raw_data
```

## Term mapping
 
Map the data to [Darwin Core Taxon](http://rs.gbif.org/core/dwc_taxon_2015-04-24.xml).
 
### language

```{r}
taxon %<>% mutate(language = "en")
```

### license

```{r}
taxon %<>% mutate(license = "http://creativecommons.org/publicdomain/zero/1.0/")
```

### rightsHolder

```{r}
taxon %<>% mutate(rightsHolder = "INBO")

```

### accessRights

```{r}
taxon %<>% mutate(accessRights = "http://www.inbo.be/en/norms-for-data-use")
```

### datasetID

```{r}
taxon %<>% mutate(datasetID = "https://doi.org/10.15468/xvuzfh")
```

### datasetName

```{r}
taxon %<>% mutate(datasetName = "Checklist of non-native freshwater fishes in Flanders, Belgium")
```

### taxonID

```{r}
taxon %<>% mutate(taxonID = raw_taxon_id)
```

### scientificName

```{r}
taxon %<>% mutate(scientificName = raw_latin_name)
```

Verify that `scientificName` contains unique values:

```{r}
any(duplicated(taxon$scientificName)) # Should be FALSE
```

### kingdom

```{r}
taxon %<>% mutate(kingdom = "Animalia")
```

### taxonRank

All taxa are species:

```{r}
taxon %<>% mutate(taxonRank = "species")
```

### nomenclaturalCode

```{r}
taxon %<>% mutate(nomenclaturalCode = "ICZN")
```

## Post-processing

Remove the original columns:

```{r}
taxon %<>% select(-starts_with("raw_"))
```

Preview data:

```{r}
taxon %>% head()
```

Save to CSV:

```{r}
write.csv(taxon, file = dwc_taxon_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create vernacular names extension

```{r start_vernacular_names}
vernacular_names <- raw_data
```

## Term mapping

Map the data to [Vernacular Names](http://rs.gbif.org/extension/gbif/1.0/vernacularname.xml).

### taxonID

```{r}
vernacular_names %<>% mutate(taxonID = raw_taxon_id)
```

### vernacularName

Vernacular names are available in two languages: English (`raw_common_name`) and Dutch (`raw_nederlandse_naam`). We will gather these columns to generate a single column containing the vernacular name (`vernacularName`) and an additional column with the language (`language`):

```{r}
vernacular_names %<>%
  gather(key = language, value = vernacularName, raw_common_name, raw_nederlandse_naam, na.rm = TRUE, convert = TRUE) %>%
  select(-language, everything()) # Move language column to the end
```

```{r}
vernacular_names %<>% arrange(taxonID)
```

### language

This column currently contains the original column name, which we will recode to the ISO 639-1 language code:

```{r}
vernacular_names %<>% mutate(language = recode(language,
  "raw_common_name" = "en",
  "raw_nederlandse_naam" = "nl"
))
```

## Post-processing

Remove the original columns:

```{r}
vernacular_names %<>% select(-starts_with("raw_"))
```

Preview data:

```{r}
vernacular_names %>% head()
```

Save to CSV:

```{r}
write.csv(vernacular_names, file = dwc_vernacular_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create distribution extension

```{r start_distribution}
distribution <- raw_data
```

## Term mapping

Map the data to [Species Distribution](http://rs.gbif.org/extension/gbif/1.0/distribution.xml).

### taxonID

```{r}
distribution %<>% mutate(taxonID = raw_taxon_id)
```

### locationID

```{r}
distribution %<>% mutate(locationID = "ISO_3166-2:BE-VLG")
```

### locality

```{r}
distribution %<>% mutate(locality = "Flemish Region")
```

### countryCode

```{r}
distribution %<>% mutate(countryCode = "BE")
```

### occurrenceStatus

```{r}
distribution %<>% mutate(occurrenceStatus = "present")
```

### establishmentMeans

We use the [GBIF controlled vocabulary](http://rs.gbif.org/vocabulary/gbif/establishment_means.xml) for this term. For this dataset, all species are `introduced` (= alien):

```{r}
distribution %<>% mutate(establishmentMeans = "introduced")
```

### eventDate

The distribution information applies to a certain date range, which we will express here as an ISO 8601 date `yyyy/yyyy` (`start_year/end_year`).

The `start_year` is the year of first observation:

```{r}
distribution %<>% mutate(start_year = raw_introduction)
```

Some `start_year` values are quite broad. We change these values to the first year for which an observation is plausible:

```{r}
distribution %<>% mutate(start_year = recode(start_year,
  "20xx" = "2000",
  "17th c." = "1601",
  "1980s" = "1980",
  "13th c." = "1201"
))
```

As there is no `end_year` information, we'll consider the publication year of Verreycken et al. (2018) as the date when the presence of the species was last verified (even though all species are considered established and will probably be present after this year as well):

```{r}
distribution %<>% mutate(end_year = "2018")
```

Create `eventDate` by combining `start_year/end_year`:

```{r}
distribution %<>% mutate(eventDate = paste(start_year, end_year, sep = "/"))
```

Compare formatted dates with original dates in `raw_introduction`:

```{r}
distribution %>%
  distinct(raw_introduction, eventDate) %>%
  arrange(raw_introduction)
```

## Post-processing

Remove the original columns:

```{r}
distribution %<>% select(-starts_with("raw_"), -start_year, -end_year)
```

Preview data:

```{r}
distribution %>% head()
```

Save to CSV:

```{r}
write.csv(distribution, file = dwc_distribution_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```

# Create description extension

In the description extension we want to include several important characteristics about the species.

## Native range

Native range information (e.g. `AS` for Asia) can be found in `raw_origin`. We create a separate dataframe to process this information:

```{r start_native_range}
native_range <- raw_data
```

Show unique values:

```{r}
native_range %>%
  distinct(raw_origin) %>%
  arrange(raw_origin)
```

`raw_origin` contains multiple values (currently not more than 2), so we separate it on `" or "` in 2 columns:

```{r}
native_range %<>% separate(raw_origin,
  into = c("native_range_1", "native_range_2"),
  sep = " or ",
  remove = FALSE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
native_range %<>% gather(
  key, value,
  native_range_1, native_range_2,
  na.rm = TRUE, # Also removes records for which there is no native_range_1
  convert = FALSE
)
```

Map values:

```{r}
native_range %<>% mutate(mapped_value = recode(value,
  "AS" = "Asia",
  "EE" = "Eastern Europe",
  "AFR" = "Africa"))
```

Show mapped values:

```{r}
native_range %>%
  select(value, mapped_value) %>%
  group_by_all() %>%
  summarize(records = n()) %>%
  arrange(value) 
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
native_range %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
native_range %<>% mutate(type = "native range")
```

## Pathway of introduction

Pathway information can be found in `raw_pathway_s`. We create a separate dataframe to process this information:

```{r start_pathway}
pathway <- raw_data
```

Show unique values:

```{r}
native_range %>%
  distinct(raw_pathway_s) %>%
  arrange(raw_pathway_s)
```

`raw_pathway_s` contains multiple values (currently not more than 2), so we separate it in 2 columns:

```{r}
pathway %<>% separate(
  raw_pathway_s,
  into = c("pathway_1", "pathway_2"),
  sep = ", ",
  remove = FALSE,
  convert = FALSE,
  extra = "merge",
  fill = "right"
)
```

Gather in a `key` and `value` column:

```{r}
pathway %<>% gather(
  key, value,
  pathway_1, pathway_2,
  na.rm = TRUE, # Also removes records for which there is no pathway_1
  convert = FALSE
)
```

We use the [CBD 2014 pathway vocabulary](https://www.cbd.int/doc/meetings/sbstta/sbstta-18/official/sbstta-18-09-add1-en.pdf) to standardize this information. The vocubulary has [these values](https://github.com/trias-project/vocab/tree/master/vocabulary/pathway).

This is a manually created overview of the abbreviations used in this dataset, their interpretation and the mapping to the CBD vocabulary:

```{r, echo = FALSE}
as.data.frame(matrix(data = c(
  "AM", "Active Migration", "corridor_water",
  "AN", "Angling/bait fish","escape_food_bait",
  "AQ", "Aquaculture", "escape_aquaculture",
  "BC", "Biological Control", "release_biologica_control",
  "BW", "Ballast Water", "stowaway_ballast_water",
  "OR", "Ornamental", "escape_ornamental",
  "UN", "Unintentional", "Unintentional"),
  nrow = 7, ncol = 3, byrow = TRUE,
  dimnames = list(c(1:7), c("abbreviation", "full name", "CBD mapping"))
))
```

Map values:

```{r}
pathway %<>% mutate(mapped_value = recode(value,
  "AM" = "corridor_water",
  "AN" = "escape_food_bait",
  "AQ" = "escape_aquaculture",
  "BC" = "release_biological_control",
  "BW" = "stowaway_ballast_water",
  "OR" = "escape_ornamental",
  "UN" = "unintentional" # Not part of CBD
))
```

Add the prefix `cbd_2014_pathway:` to refer to this standard, except for `unintentional`, which has no alternative in the standard:

```{r}
pathway %<>% mutate(mapped_value = case_when (
  mapped_value != "unintentional" ~ paste("cbd_2014_pathway", mapped_value, sep = ":"),
  mapped_value == "unintentional" ~ mapped_value))
```

Show mapped values:

```{r}
pathway %>%
  select(value, mapped_value) %>%
  group_by_all() %>%
  summarize(records = n()) %>%
  arrange(value) 
```

Drop the `key` and `value` columns and rename `mapped_value` as `description`:

```{r}
pathway %<>%
  select(-key, -value) %>%
  rename(description = mapped_value)
```

Create a `type` field to indicate the type of description:

```{r}
pathway %<>% mutate(type = "pathway")
```

## Invasion stage

Invasion stage information (e.g. `casual`) can be found in `raw_status`. We create a separate dataframe to process this information:

```{r start_invasion_stage}
invasion_stage <- raw_data
```

Show unique values:

```{r}
native_range %>%
  distinct(raw_status) %>%
  arrange(raw_status)
```

We use the [invasion stage vocabulary from Blackburn et al. (2011)](http://doc.rero.ch/record/24725/files/bach_puf.pdf) to standardize this information.

* `A` and `A*` in the raw data file stand for `acclimatized` and refers to the status where the alien species has overcome the geographical barrier, the captivity/cultivation barrier and the survival barier, but where it fails to establish because individuals in the population fail to reproduce. This includes species (like Nile tilapia) in secluded warmer water of power plants. In Blackburn et al. (2011), this corresponds to the term `casual`.
* `N` in the raw data file stands for `naturalized`. We decided **not to use** this term, because often, there's no sensible criterium to distinguish between casual/naturalised and naturalised/established. Thus, here, we consider all `naturalized` species to be `established`.

```{r}
invasion_stage %<>% mutate(description = case_when(
  raw_status == "A" ~ "casual",
  raw_status == "A*" ~ "casual",
  raw_status == "N" ~ "established"
))
```

Create a `type` field to indicate the type of description:

```{r}
invasion_stage %<>% mutate(type = "invasion stage")
```

## Union descriptions

Union native range, pathway of introduction and invasion stage:

```{r start_description_ext}
description_ext <- bind_rows(native_range, pathway, invasion_stage)
```

## Term mapping

Map the data to [Taxon Description](http://rs.gbif.org/extension/gbif/1.0/description.xml).

### taxonID

```{r}
description_ext %<>% mutate(taxonID = raw_taxon_id)
```

### description

```{r}
description_ext %<>% mutate(description = description)
```

### type

```{r}
description_ext %<>% mutate(type = type)
```

### language

```{r}
description_ext %<>% mutate(language = "en")
```

## Post-processing

Remove the original columns:

```{r}
description_ext %<>% select(-starts_with("raw_"))
```

Move `taxonID` to the first position:

```{r}
description_ext %<>% select(taxonID, everything())
```

Sort on `taxonID` to group description information per taxon:

```{r}
description_ext %<>% arrange(taxonID)
```

Preview data:

```{r}
description_ext %>% head(10)
```

Save to CSV:

```{r}
write.csv(description_ext, file = dwc_description_file, na = "", row.names = FALSE, fileEncoding = "UTF-8")
```
